---
title: "Basics of Panel Data"
subtitle: "<html><div style='float:left'></div><hr color='#EB811B' size=1px width=0px></html>"
author: Ian McCarthy | Emory University
date: Workshop on Causal Inference with Panel Data
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code {
  font-size: 15px;
}
.remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
knitr::opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T,# echo=F, warning=F, message=F
  warning = FALSE, 
  message = FALSE, 
  cache.lazy = FALSE,
  error=TRUE
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, fixest, readstata13)
set.seed(1234)
```

# Table of contents

1. [What are Panel Data](#panel)
2. [Estimation](#estimation)
3. [In Practice](#handson)
4. [Semantics](#semantics)

---
class: inverse, center, middle
name: panel

# What are Panel Data?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>


---
# Nature of the Data

- Repeated observations of the same units over time


--
**Notation**
- Unit $i=1,...,N$ over several periods $t=1,...,T$, which we denote $y_{it}$
- Treatment status $D_{it}$
- Regression model, <br>
$y_{it} = \delta D_{it} + u_{i} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$

---
# Benefits of Panel Data

- *May* overcome certain forms of omitted variable bias
- Allows for unobserved but time-invariant factor, $u_{i}$, that affects both treatment and outcomes


--
**Still assumes**
- No time-varying confounders 
- Past outcomes do not directly affect current outcomes
- Past outcomes do not affect treatment (reverse causality)

---
# Some textbook settings

- Unobserved "ability" when studying schooling and wages
- Unobserved "quality" when studying physicians or hospitals

---
class: inverse, center, middle
name: estimation

# Estimating Regressions with Panel Data

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>

---
# Regression model

$y_{it} = \delta D_{it} + u_{i} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$

---
# Fixed Effects

$y_{it} = \delta D_{it} + u_{i} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$


--
- Allows correlation between $u_{i}$ and $D_{it}$
- Physically estimate $u_{i}$ in some cases via set of dummy variables
- More generally, "remove" $u_{i}$ via:
  - "within" estimator
  - first-difference estimator
  
---
# Within Estimator
$y_{it} = \delta D_{it} + u_{i} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$


--
- Most common approach (default in most statistical software)
- Equivalent to demeaned model,<br>
$y_{it} - \bar{y}_{i} = \delta (D_{it} - \bar{D}_{i}) + (u_{i} - \bar{u}_{i}) + (\epsilon_{it} - \bar{\epsilon}_{i})$
- $u_{i} - \bar{u}_{i} = 0$ since $u_{i}$ is time-invariant
- Requires *strict exogeneity* assumption (error is uncorrelated with $D_{it}$ for all time periods)

---
# First-difference
$y_{it} = \delta D_{it} + u_{i} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$


--
- Instead of subtracting the mean, subtract the prior period values<br>
$y_{it} - y_{i,t-1} = \delta(D_{it} - D_{i,t-1}) + (u_{i} - u_{i}) + (\epsilon_{it} - \epsilon_{i,t-1})$
- Requires exogeneity of $\epsilon_{it}$ and $D_{it}$ only for time $t$ and $t-1$ (weaker assumption than within estimator)
- Sometimes useful to estimate both FE and FD just as a check


---
# Keep in mind...

- Discussion only applies to linear case or very specific nonlinear models
- Fixed effects can't solve reverse causality
- Fixed effects doesn't address unobserved, time-varying confounders
- Can't estimate effects on time-invariant variables
- May "absorb" a lot of the variation for variables that don't change much over time

---
class: inverse, center, middle
name: handson

# Seeing things in action

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>


---
# Within Estimator (Default)

.pull-left[
**Stata**<br>
```{stata, eval=FALSE}
ssc install bcuse
bcuse wagepan
tsset nr year
xtreg lwage exper expersq, fe
```
]

.pull-right[
**R**<br>
```{r, eval=FALSE}
library(readstata13)
library(fixest)
wagepan <- read.dta13("http://fmwww.bc.edu/ec-p/data/wooldridge/wagepan.dta")
feols(lwage~exper + expersq | nr, data=wagepan)
```
]


---
# Within Estimator (Manually Demean)

.pull-left[
**Stata**<br>
```{stata, eval=FALSE}
ssc install bcuse
bcuse wagepan
foreach x of varlist lwage exper expersq {
  egen mean_`x'=mean(`x')
  egen demean_`x'=`x'-mean_`x'
}
reg demean_lwage demean_exper demean_expersq
```
]

.pull-right[
**R**<br>
```{r, eval=FALSE}
library(readstata13)
wagepan <- read.dta13("http://fmwww.bc.edu/ec-p/data/wooldridge/wagepan.dta")
wagepan <- wagepan %>%
  group_by(nr) %>%
  mutate(demean_lwage=lwage - mean(lwage),
         demean_exper=exper - mean(exper),
         demean_expersq=expersq - mean(expersq))
summary(lm(demean_lwage~demean_exper + demean_expersq, data=wagepan))
```
]


---
# First differencing

.pull-left[
**Stata**<br>
```{stata, eval=FALSE}
ssc install bcuse
bcuse wagepan
reg d.lwage d.exper d.expersq, noconstant
```
]

.pull-right[
**R**<br>
```{r, eval=FALSE}
library(readstata13)
wagepan <- read.dta13("http://fmwww.bc.edu/ec-p/data/wooldridge/wagepan.dta")
wagepan <- wagepan %>%
  group_by(nr) %>%
  arrange(year) %>%
  mutate(fd_lwage=lwage - lag(lwage),
         fd_exper=exper - lag(exper),
         fd_expersq=expersq - lag(expersq)) %>%
  na.omit()
summary(lm(fd_lwage~0 + fd_exper + fd_expersq, data=wagepan))
```
]


---
class: inverse, center, middle
name: semantics

# Some Communication Problems!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1055px></html>
